---
title: "ESM 244 Assignment 1 Task 2"
author: "Sydney Rilum"
date: "1/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(janitor)
library(here)


library(lubridate) # For easier dates & times
library(gghighlight) # For useful highlighting
library(plotly) # Interactive graphs
library(patchwork) # For compound figures


# For Part 1 (PCA)
library(ggfortify) # For PCA biplot

```



## Introduction 
1) A useful descriptive introductory summary (3 - 4 sentences) that helps the audience understand the data (include a citation as necessary) and what you’ll be exploring

This report provides an exploratory ....
Principal components analysis is an ordination method allowing us to glean as much about our multivariate data as possible in a simplified number of dimensions. 

explore variable relationships and clustering by species in a PCA biplot.


## Principal Components Analysis (PCA)

2) All of your organized and well-annotated code (with warnings/messages suppressed) you wrote to wrangle data then run PCA, and to create a professional looking PCA biplot that appears (with a caption) in your knitted HTML

```{r}
# Read in .csv file and clean up data
world_envi <- read_csv(here("data", "world_env_vars.csv")) %>% 
  clean_names()

```


```{r}
# PCA
world_pca <- world_envi %>% 
  select(body_mass_g, ends_with("_mm")) %>%  # keeps variables in mm units
  drop_na() %>% 
  scale() %>%  # values scaled
  prcomp() # makes data frame into a list of info

# Create a dataset that drops NAs but contains all 8 variables to add aesthetics to biplot, matches PCA dataset
world_complete <- world_envi %>% 
  drop_na(body_mass_g, ends_with("_mm"))


# Create a biplot using autoplot (from ggfortify package)
autoplot(world_pca,
         data = world_complete,
         colour = 'species',
         loadings = TRUE,
         loadings.label = TRUE) + # shows loadings of each variable
  theme_minimal() # can customize just like ggplot


# autoplot(penguin_pca,
#         data = penguin_complete,
#         loadings = TRUE,
 #        colour = 'species',
  #       loadings.label = TRUE,
   #      loadings.colour = "black",
    #     loadings.label.colour = "black",
     #    loadings.label.vjust = -0.5
     #    ) +
#  scale_color_manual(values = c("blue","purple","orange")) +
 # scale_fill_manual(values = c("blue","purple","orange")) +
  #theme_minimal()

# Variance explained by each PC
screeplot(world_pca, type = "lines")

# See the loadings (weighting for each principal component)
world_pca$rotation

```



## Summary
3) A brief summary (nicely formatted and professionally written bulletpoints are fine) highlighting some major takeaways from your PCA that can be gleaned from the biplot



