---
title: "Sierra Lake Mountain Yellow-Legged Frog Count Visualizations"
author: "Sydney Rilum"
date: "1/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(janitor)
library(here)
library(readxl)
library(lubridate) # for easier dates & times
library(patchwork) # for creating compound figures
```

```{r}
# Read in .xlsx file and clean up data
frogs <- read_excel(here("data","sierra_amphibians.xlsx")) %>% 
  mutate(across(where(is.character), tolower)) # convert all character observations to lowercase 

frogs$lake_id <- as.character(as.numeric(frogs$lake_id)) # converts lake_id from a numeric to character class
```

```{r}
# Create a subset of data for mountain yellow-legged frog (Rana muscosa) counts each year across all water bodies, by life stage excluding the ‘EggMass’ level. 
yellow_frog_count <- frogs %>% 
  select(survey_date, amphibian_species, amphibian_life_stage, amphibian_number) %>% 
  filter(amphibian_species == "ramu") %>%   # filter for RAMU (Rana muscosa) species
  filter(amphibian_life_stage != "eggmass") %>%   # remove eggmass life stage
  mutate(year = year(survey_date)) %>%   # make new column for year only
  group_by(year, amphibian_life_stage) %>%   # group by year and life stage
  summarize(total_count = sum(amphibian_number, na.rm = TRUE))  # add up total counts for each life stage each year


# Create a bar graph of total mountain yellow-legged frog counts each year by life stage 
graph_a <- ggplot(data = yellow_frog_count, aes(x = year, y = total_count)) +
  geom_col(aes(fill = amphibian_life_stage)) +
  facet_wrap(~ amphibian_life_stage) +
  scale_color_manual(aes(amphibian_life_stage = c("darkorange","springgreen4", "cornflowerblue"))) +
  labs(x = "Year",
       y = "Total Count",
       title = "Total Mountain Yellow-Legged Frog Counts by Life Stage") 
  
```

```{r}
# Create a new subset of data of total counts (over all years of the study) of combined adult and subadult mountain yellow-legged frogs (Rana muscosa) observed in the 5 lakes with the greatest total observed counts. 
yellow_frog_count_lakes <- frogs %>% 
  select(lake_id, amphibian_species, amphibian_life_stage, amphibian_number, amphibian_location) %>% 
  filter(amphibian_location == "in lake") %>%  # filter for frogs observed in lake only
  filter(amphibian_species == "ramu") %>%   # filter for RAMU (Rana muscosa) species
  filter(amphibian_life_stage %in% c("adult", "subadult")) %>%  # filter for adult and subadult frogs only
  group_by(lake_id) %>%   # group by lake ID
  summarize(count = sum(amphibian_number, na.rm = TRUE)) %>%  # counts total number of frogs observed at each lake(id)
  mutate(lake_id = fct_reorder(lake_id, count)) %>%  # order lake_ids by count (largest to smallest)
  slice_max(count, n = 5)  # only keeps top 5 lake_ids with the largest count

# Create a column graph containing total counts (over all years of the study) of combined adult and subadult endangered mountain yellow-legged frogs (Rana muscosa) observed in the 5 lakes with the greatest total observed counts. 
graph_b <- ggplot(data = yellow_frog_count_lakes, aes(y = lake_id, x = count)) +
  geom_col(fill = "skyblue4") +
  labs(x = "Total adult and subadult counts",
       y = "Lake ID",
       title = "Top 5 Lakes with Largest Total Mountain Yellow-Legged Frog Count")

```

```{r}
# Combine graphs a and b into a single compound figure using the `patchwork` package
graph_a / graph_b & theme_minimal() +
  theme(legend.position = "none")

```

**Figure 1.** The top bar graph displays the total mountain yellow-legged frog (*Rana muscosa*) count across all water bodies, by life stage (adult, subadult, and tadpole) and for each year in the study (1995-1997, 2000-2002). The bottom column graph illustrates the top five lakes with the largest total mountain yellow-legged frog count (excluding frogs in the tadpole life stage) over all years of the study. Data: Knapp et al. (2020).



**Data Citation:** [Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative.](https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3)
